Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-02-06 TEBenchmarkPaper/article.tex
Fix citation
Use Mathematical Theory Exploration
Reword, since we're not investigating e.g. concept formation

\emph{Automated theory exploration} (ATE), also known as
\emph{conjecture generation/formation}, is the open-ended problem of producing
conjectures about a given logical theory which are somehow ``interesting''.

@@

Differences in background and motivation continue to be relevant
today. While the majority of work in automating mathematics has been
in symbolic manipulation and theorem proving, we are concerned here
with other aspects of mathematics, including the construction of
concepts, axioms, conjectures, examples and theorems. This is
varyingly known as ``Automated Theory Formation''
\cite{lenat:77,colton:book}, ``Automated Mathematical Theory
Exploration'', ``Mathematical Theory Exploration''
\cite{buchberger:06} (also sometimes prefaced with ``Computer-Aided''
or ``Algorithm-Supported''), ``Automated Mathematical Discovery''
\cite{epstein:91,colton:interestingness,esarm2008}, ``Concept
Formation in Discovery Systems'' \cite{haase}, and ``Automated Theorem
Discovery'' \cite{roy}. Such a plethora of terminology can be
unhelpful and can mask similarities between the different fields. In
particular, the twin strands of Automated Theory Formation and
Automated Mathematical Theory Exploration seem to be developing
somewhat independently without a clear differentiating
methodology. Below we discuss commonalities and differences between
the two schools of thought.

@@

Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical development
\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly proposed
metric. Methods for finding out what that means have come from neuro-scientists,
such as the study by Zeki et al in which they test whether mathematicians'
experiences of abstract beauty correlates with the same brain activity as
experiences of sensory beauty \cite{zeki}; computer scientists, such as Colton's
metrics above which are proposed based largely on a computer scientist's
``intuition'' and argumentation about why a metric would be important (and
previous use of such metrics, and - in one isolated case (surprisingingness) -
on a single quote from a mathematician\footnote{Similarly, classification of a
MATHsAiD result as either interesting or uninteresting \cite{roy} was made by
the main system developer: in both cases the main developer had a background in
mathematics.}; and mathematicians themselves, such as Gowers' suggestion that we
can identify features which are commonly associated with good proofs
\cite{gowers}. All of these approaches rest upon the assumption that it makes
sense to speak of ``the good'' in mathematics. This assumption is supported by
concepts such as Erdos's ``The Book'' -- a theoretical construct of a book which
contains the most elegant proof of every theorem (brought to life in
\cite{Martin Aigner and Günter M. Ziegler Proofs from THE BOOK}). However,
empirical studies by psychologists such as Inglis call into question such
assumptions: work by Inglis and colleagues has shown that there is not a single
standard of validity among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
nor beauty. %FIXME needs citation

>>>>

\emph{Mathematical theory exploration} (MTE), also known as
\emph{conjecture generation/formation}, is the open-ended problem of producing
conjectures about a given logical theory which are somehow ``interesting''.

@@

Differences in background and motivation continue to be relevant
today. While the majority of work in automating mathematics has been
in symbolic manipulation and theorem proving, we are concerned here
with other aspects of mathematics, including the construction of conjectures,
(counter-)examples and theorems. These tasks, along with the formation of new
definitions and concepts, are varyingly known as
``Automated Theory Formation''~\cite{lenat:77,colton:book},
``Automated Mathematical Theory Exploration'',
``Mathematical Theory Exploration''~\cite{buchberger:06} (also sometimes
prefaced with ``Computer-Aided'' or ``Algorithm-Supported''),
``Automated Mathematical Discovery''~\cite{epstein:91,colton:interestingness,esarm2008},
``Concept Formation in Discovery Systems''~\cite{haase}, and
``Automated Theorem Discovery''~\cite{roy}. Such a plethora of terminology can
be unhelpful and can mask similarities between the different fields. In
particular, the twin strands of Automated Theory Formation and
Automated Mathematical Theory Exploration seem to be developing
somewhat independently without a clear differentiating
methodology. Below we discuss commonalities and differences between
the two schools of thought.

@@

Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical development
\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly proposed
metric. Methods for finding out what that means have come from neuro-scientists,
such as the study by Zeki et al in which they test whether mathematicians'
experiences of abstract beauty correlates with the same brain activity as
experiences of sensory beauty \cite{zeki}; computer scientists, such as Colton's
metrics above which are proposed based largely on a computer scientist's
``intuition'' and argumentation about why a metric would be important (and
previous use of such metrics, and - in one isolated case (surprisingingness) -
on a single quote from a mathematician\footnote{Similarly, classification of a
MATHsAiD result as either interesting or uninteresting \cite{roy} was made by
the main system developer: in both cases the main developer had a background in
mathematics.}; and mathematicians themselves, such as Gowers' suggestion that we
can identify features which are commonly associated with good proofs
\cite{gowers}. All of these approaches rest upon the assumption that it makes
sense to speak of ``the good'' in mathematics. This assumption is supported by
concepts such as Erdos's ``The Book'' -- a theoretical construct of a book which
contains the most elegant proof of every theorem (brought to life in
\cite{aigner2010proofs}). However,
empirical studies by psychologists such as Inglis call into question such
assumptions: work by Inglis and colleagues has shown that there is not a single
standard of validity among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
nor beauty. %FIXME needs citation
